Predicting Kamala Harris's Victory in the 2024 US Election

### 2024-10-31 paper/paper.qmd
Paper updated.

---
title: "Predicting Kamala Harris's Victory in the 2024 US Election"
subtitle: "An Investigation into Poll Data Using Multiple Linear Regression"
author: 
  - Xuanle Zhou
  - Yongqi Liu
  - Yuxuan Wei
thanks: "Code and data are available at: https://github.com/wyx827/2024USpresidentialelection.git."
date: today
date-format: long
toc: true
abstract: "The study predicting the 2024 U.S. presidential election is important as it provides the predicted outcomes for the US election. Analysis is based on state-by-state electoral votes, and these outcomes can influence campaign strategies and shape public discourse. The results indicate Kamala Harris will win by over 270 votes in this election cycle."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# Install the required packages
#install.packages(c("tidyverse", "ggplot2", "rstanarm", "knitr", "usmap", "arrow"))
library(tidyverse)
library(ggplot2)
library(rstanarm)
library(knitr)
library(usmap)
library(arrow)

analysis_data <- read_csv(here::here("data/02-analysis_data/cleaned_US_voting.csv"))
```

@@

@sec-data expains the data we used to build up the prediction mode.

@@

## Outcome variables

@@

## Predictor variables

@@

### End Date

The end date indicates when each poll concluded. @fig-end-date shows that as the final result of the U.S. election approached, more polls were conducted and completed. This trend reflects an increased interest in capturing public opinion and predicting electoral outcomes as election day neared, emphasizing the significance of real-time sentiment analysis during this critical period.

```{r}
#| label: fig-end-date
#| fig-cap: Distribution of End Dates
#| echo: false
#| warning: false
#| message: false

ggplot(analysis_data, aes(x = as.Date(end_date))) + 
  geom_line(stat = "count", color = "black", size = 1) + 
  labs(title = "Distribution of End Dates", x = "End Date", y = "Count") + 
  theme_minimal()

# Model

To model Donald Trump’s polling percentages over time, we employed a multiple linear regression framework. This approach estimates the relationship between polling percentages and various predictors by fitting a linear equation to the data. By analyzing the coefficients, we can quantify the impact of each predictor on Trump’s polling percentages, while also assessing the overall fit of the model and making predictions based on the observed trends.

Here we briefly describe the multiple linear regression model used to investigate the winning probability of Trump. Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

- Numeric Grade (numeric_grade): Reflects the quality rating of the pollster.
- Sample Size (sample_size): The number of respondents in the poll.
- State (state): A categorical variable for different U.S. states.
- Transparency Score (transparency_score): A measure of how transparent the polling data and methodology are.
- End Date (end_date): The date the poll was completed, which might capture trends over time.

@@

\begin{align}
\text{pct}_i &= \beta_0 + \beta_1 \cdot \text{numeric\_grade}_i + \beta_2 \cdot \text{transparency\_score}_i \\
             &\quad + \beta_3 \cdot \text{sample\_size}_i + \beta_4 \cdot \text{state}_i + \beta_5 \cdot \text{end\_date}_i + \epsilon_i \\
\epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align}

@@

- Intercept ($\beta_0$): This is the predicted Trump polling percentage when all predictors (numeric grade, sample size, state, transparency score, and end date) are at their baseline or zero value.
- Numeric Grade ($\beta_1$): This coefficient measures how much Trump’s polling percentage changes as the pollster’s numeric grade increases. A positive and significant coefficient would indicate that higher-rated pollsters report better polling numbers for Trump, while a negative coefficient would suggest the opposite.
- Sample Size ($\beta_2$): This measures the impact of the number of respondents on Trump’s polling percentage. A positive coefficient would indicate that larger sample sizes are associated with higher polling percentages for Trump.
- State ($\beta_3$): The coefficients for the state variable represent differences in Trump’s polling percentage in each state compared to the reference state (baseline category). For example, if the coefficient for Florida is negative, it means Trump polls lower in Florida compared to the reference state.
- Transparency Score ($\beta_4$): This coefficient shows how much Trump’s polling percentage is affected by the transparency of the poll. A positive coefficient would indicate that polls with higher transparency tend to report higher polling percentages for Trump, whereas a negative coefficient would imply the opposite.
- End Date ($\beta_5$): The end date is a time-related variable, capturing trends over time. A positive and significant coefficient would suggest that Trump’s polling percentage has increased as the election date approaches, while a negative coefficient would suggest a decrease in his polling numbers over time.

## Interpretation

The posterior distributions of the parameters allow us to quantify the uncertainty around each effect:\
- The coefficient for end_date informs us about how Trump’s polling percentages have evolved over time. A positive coefficient would suggest an upward trend, while a negative coefficient would indicate a decline.\
- The coefficient for numeric_grade captures the impact of pollster quality on the polling percentage. High-quality pollsters may produce different estimates compared to lower-quality ones.\
- The state-level effects account for regional differences in Trump’s support. Some states may show significantly higher or lower levels of support, even after adjusting for the time of the poll and pollster quality.

## Model Justification
@tbl-summary shows the summary table for the model. And then evaluate it on the test set. It appears as though the model is having difficulty identifying Trump supporters. The reson why the multiple regression is applied is because.....

```{r}
#| echno: false
#| label: tbl-summary
#| tbl-cap: Regression Model Estimate Summary
#| echo: false
#| warning: false
#| message: false

regression_model<-readRDS(here::here("models/regression_model.rds"))
summary<-summary(regression_model)

# Extract the coefficients to form a table
coefficients_table <- summary$coefficients

# Display the first 10 rows of the coefficients table using kable, formatted to 4 decimal places
kable(head(coefficients_table, 10), digits = 2)
```

# Results

@@

Electoral Vote Count:
- Trump Electoral Votes: 78
- Harris Electoral Votes: 416
These results indicate that based on the current model predictions, Harris is expected to win the 2024 U.S. Presidential Election, securing 416 electoral votes, compared to Trump’s 78 electoral votes. The election outcome hinges on several battleground states, where the vote margins are predicted to be narrow.

```{r tbl-prediction}
#| echno: false
#| label: tbl-prediction
#| tbl-cap: Prediction for Trump and Harris by Electoral College
#| echo: false
#| warning: false
#| message: false

# Read the test data from the .parquet file
analysis_data_test <- read_parquet(here::here("data/02-analysis_data/test_data.parquet"))
analysis_data_train<- read_parquet(here::here("data/02-analysis_data/train_data.parquet"))
# Filter out states in the test data that don't exist in the training data

analysis_data_test <- analysis_data_test %>%
 filter(state %in% unique(analysis_data_train$state))

@@

# Create the table using kable
kable(state_predictions[, c("state", "predicted_percent_trump", "electoral_votes", "winner")], 
      col.names = c("State", "Trump Predicted %", "Electoral Votes", "Winner"), 
      digits = 2)

```{r}
#| echno: false
#| label: tbl-missing
#| tbl-cap: Missing States with Predicted Winner Based on Historical Lean
#| echo: false
#| warning: false
#| message: false
# Create the data frame with missing states, electoral votes, and predicted winner based on historical lean
missing_states <- data.frame(
  state = c("Alabama", "Alaska", "Arkansas", "Connecticut", "Delaware", "Hawaii", "Idaho", 
            "Illinois", "Indiana", "Kansas", "Kentucky", "Louisiana", "Mississippi", "North Dakota", 
            "Oklahoma", "Oregon", "South Carolina", "Tennessee", "Utah", "Washington", 
            "West Virginia", "Wyoming", "District of Columbia"),
  Electoral_Votes = c(9, 3, 6, 7, 3, 4, 4, 19, 11, 6, 8, 8, 6, 3, 7, 8, 9, 11, 6, 12, 4, 3, 3),
  Prediction_Based_on_Historical_Lean = c("Trump", "Trump", "Trump", "Harris", "Harris", 
                                          "Harris", "Trump", "Harris", "Trump", "Trump", 
                                          "Trump", "Trump", "Trump", "Trump", "Trump", 
                                          "Harris", "Trump", "Trump", "Trump", "Harris", 
                                          "Trump", "Trump", "Harris")
)

# Display the table with kable
kable(missing_states, col.names = c("State", "Electoral Votes", "Prediction Based on Historical Lean"))

## Predicted Electoral Outcomes by State
The following map (@fig-statemap) shows the predicted winner for each state in the 2024 U.S. Presidential Election, based on the regression model's predicted vote percentages for Trump and Harris.

```{r tbl-prediction}
#| echno: false
#| label: fig-statemap
#| tbl-cap: Predicted Winner of 2024 US Presidential Election by State
#| echo: false
#| warning: false
#| message: false
state_predictions <- merge(state_predictions, missing_states, by = "state", all = TRUE)

@@

```
The predicted outcome in the regression model reflects the geographic voting patterns, with Trump winning in traditionally Republican-leaning states like Alabama, Missouri, and Wyoming, while Harris dominates in Democratic strongholds such as California, New York, and Illinois. However, key battleground states such as Florida and Arizona are predicted to favor Harris, potentially determining the overall election outcome.

```{r}
#| echno: false
#| label: fig-statenum
#| tbl-cap: Numerical Predicted Winner of 2024 US Presidential Election by State
#| echo: false
#| warning: false
#| message: false

@@

# Display the summary table in kable format
cat("\nTotal Electoral Votes:\n")
kable(summary_df, col.names = c("Candidate", "Total Electoral Votes"), format = "markdown")
```

@@

## What Is Done in This Paper?
This paper uses a regression model to predict the outcome of the 2024 U.S. Presidential Election between Donald Trump and Kamala Harris. By analyzing historical voting data, the model forecasts the percentage of votes Trump is expected to receive in each state. These predictions are combined with the state's electoral vote allocation to determine the overall winner. The findings indicate a significant electoral victory for Harris, projecting 416 electoral votes for her compared to 78 for Trump. A geographic map is also presented to visualize state-level outcomes and highlight regional voting trends.

## What Do We Learn About the World?
The study highlights the persistence of geographic voting patterns in U.S. elections. Even with changes in candidates, many states follow their historical voting tendencies. Republican-leaning states like Alabama and Wyoming are projected to support Trump, while Democratic-leaning states such as California and New York favor Harris. This pattern shows the strong influence of regional political loyalties, which continue to shape election outcomes.

## Another Thing We Learn About the World
The analysis underscores the importance of battleground states in determining the election's outcome. States like Florida and Arizona, predicted to lean toward Harris, play a decisive role in the projected electoral victory. This illustrates how close contests in a few states can influence the overall election result, emphasizing the significant impact these states have in shaping the presidency.

## What Is Left to Learn or How Should We Proceed in the Future?
Future research should incorporate more detailed data, including demographic shifts and economic factors that may influence voter preferences. It would be valuable to examine how emerging political movements and changes in communication strategies affect elections, particularly in closely contested states. Post-election analysis comparing predictions with actual results will also be essential to improving the accuracy of models and understanding how unforeseen events may alter outcomes.

## Weaknesses and Next Steps
While the model offers a useful prediction, it has limitations. It relies heavily on past voting data, which may not fully capture future changes in voter behavior or external factors that could affect the election. The model also simplifies the election by focusing on state-level percentages, without accounting for variables such as turnout, third-party candidates, or unexpected political shifts. Additionally, the assumption that past voting patterns will continue might not be accurate in a rapidly changing political landscape. The logistic regression model is limited by the lack of interaction terms between states, which may obscure regional trends or external factors that influence neighboring states. Additionally, the model assumes that polling data accurately reflects voter intentions, but biases or missing data in key regions could affect the predictions.

Future iterations of this model should incorporate voter turnout predictions, demographic variables, and real-time polling updates to better capture the dynamics of voter behavior. Including social media sentiment analysis and analyzing cross-state interactions could refine the prediction model, offering more nuanced insights into election forecasts.

@@

# Appendix {-}

@@

### **Survey Structure**:

@@

```{r}
#| echo: false
#| label: fig-diagnosis-3
#| tbl-cap: Linear Regression Assumptions hold
#| warning: false
#| message: false
#| fig.width: 10
#| fig.height: 8

par(mfrow = c(2, 2))  
plot(regression_model)

@@

## Calculate Mean Squared Error (MSE) on test data
```{r}
#| echo: false
#| label: fig-diagnosis-1
#| tbl-cap: Calculate Mean Squared Error (MSE) on test data
#| warning: false
#| message: false

predicted_values <- predict(regression_model, newdata = analysis_data_test)
actual_values <- analysis_data_test$percent

mse <- mean((predicted_values - actual_values)^2)
print(mse)
```

```{r}
#| echo: false
library(caret)
set.seed(856)
train_control <- trainControl(method = "cv", number = 10)
cv_model <- train(percent ~ numeric_grade + sample_size + state + transparency_score + end_date,
                  data = analysis_data_train,
                  method = "lm",
                  trControl = train_control)

print(cv_model$results)

```{r}
#| echo: false
baseline_prediction <- mean(analysis_data_train$percent)
baseline_mse <- mean((analysis_data_test$percent - baseline_prediction)^2)
print(baseline_mse)
```

```{r}
#| echo: false
# Calculate R-squared on test data
rss <- sum((predicted_values - actual_values)^2)  # Residual Sum of Squares
tss <- sum((actual_values - mean(actual_values))^2)  # Total Sum of Squares

r_squared <- 1 - (rss/tss)
print(r_squared)
```

```{r}
#| echo: false
# Create a correlation matrix to check for collinearity
cor_matrix <- cor(analysis_data_train[, c("numeric_grade", "sample_size", "transparency_score")], use = "complete.obs")
print(cor_matrix)
```

```{r}
#| echo: false
# Check for sparse levels in 'state'
state_counts <- table(analysis_data_train$state)
print(state_counts)
```

>>>>

---
title: "Predicting Kamala Harris's Victory in the 2024 US Election"
subtitle: "An Investigation into Poll Data Using Multiple Linear Regression"
author: 
  - Xuanle Zhou
  - Yongqi Liu
  - Yuxuan Wei
thanks: "Code and data are available at: https://github.com/wyx827/2024USpresidentialelection.git"
date: today
date-format: long
toc: true
abstract: "The study predicting the 2024 U.S. presidential election is important as it provides the predicted outcomes for the US election. Analysis is based on state-by-state electoral votes, and these outcomes can influence campaign strategies and shape public discourse. The result indicates Kamala Harris will win by over 270 votes in this election cycle."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# Install the required packages
#install.packages(c("tidyverse", "ggplot2", "rstanarm", "knitr", "usmap", "corrplot", "arrow"))
library(tidyverse)
library(ggplot2)
library(rstanarm)
library(knitr)
library(usmap)
library(corrplot)
library(arrow)
analysis_data <- read_csv(here::here("data/02-analysis_data/cleaned_US_voting.csv"))
```

@@

@sec-data explains the data we used to build up the prediction mode.

@sec-model
@sec-results
@sec-discussion
@sec-appendix

@@

## Outcome Variable

@@

## Predictor Variables

@@

### Days Until Election

The **days until election** represents the remaining time leading up to the final election day. As shown in the figure, the support percentage for Donald Trump fluctuated over time, with a slightly downward trend as the election approached, as indicated by the dashed trend line. This suggests that while public opinion varied significantly, there was a decrease in support as election day approached. This trend highlights the importance of monitoring real-time polling data, as sentiment may shift in the critical months leading up to a closely contested election.

```{r}
#| label: fig-end-date
#| fig-cap: Trend of Support Percentage for Donald Trump Over Days Until Election. Fluctuations in support levels are observed as the election approaches, with a slight downward trend indicated by the dashed curve.
#| echo: false
#| warning: false
#| message: false

ggplot(analysis_data, aes(x = days_until_election, y = percent)) + 
  geom_line(color = "skyblue", size = 1.2) +  
  geom_smooth(method = "loess", color = "pink", linetype = "dashed", size = 1) +  
  labs(x = "Days Until Election", y = "Support Percentage") + 
  theme_minimal(base_size = 14) +  
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),  #
    axis.title.x = element_text(margin = margin(t = 10)),  
    axis.title.y = element_text(margin = margin(r = 10)),  
    axis.text = element_text(color = "#555555"),  
    panel.grid.minor = element_blank()  )
```

# Model {#sec-model}

To model Donald Trump’s polling percentages over time, we used a multiple linear regression framework, which estimates the relationship between polling percentages and various predictors by fitting a linear equation to the data. Analyzing the coefficients allows us to quantify the impact of each predictor on Trump’s polling percentages, while also assessing the overall fit of the model for reliable predictions.

To prevent overfitting, we applied a train-test data split. The training data is used to build the model, enabling it to learn patterns and relationships within the data. The test data, which the model has not seen before, serves to evaluate its performance on new, unseen data. This separation ensures that the model captures patterns that will generalize beyond the training dataset, rather than just memorizing it.

Below, we briefly describe the multiple linear regression model used to examine Trump’s winning probability. Additional details and model diagnostics are provided in [Appendix -@sec-model-details].

@@

- Numeric Grade (numeric_grade): Reflects the quality rating of the pollster.
- Sample Size (sample_size): The number of respondents in the poll.
- State (state): A categorical variable for different U.S. states.
- Transparency Score (transparency_score): A measure of how transparent the polling data and methodology are.
- Days Until Election (days_until_election): The left days until the US election.

@@

\begin{align}
\text{pct}_i &= \beta_0 + \beta_1 \cdot \text{numeric\_grade}_i + \beta_2 \cdot \text{transparency\_score}_i \\
             &\quad + \beta_3 \cdot \text{sample\_size}_i + \beta_4 \cdot \text{state}_i + \beta_5 \cdot \text{days\_until\_election}_i + \epsilon_i \\
\epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align}

@@

- Intercept ($\beta_0$): This is the predicted Trump polling percentage when all predictors (numeric grade, sample size, state, transparency score, and end date) are at their baseline or zero value.
- Numeric Grade ($\beta_1$): This coefficient measures how much Trump’s polling percentage changes as the pollster’s numeric grade increases. A positive and significant coefficient would indicate that higher-rated pollsters report better polling numbers for Trump, while a negative coefficient would suggest the opposite.
- Sample Size ($\beta_2$): This measures the impact of the number of respondents on Trump’s polling percentage. A positive coefficient would indicate that larger sample sizes are associated with higher polling percentages for Trump.
- State ($\beta_3$): The coefficients for the state variable represent differences in Trump’s polling percentage in each state compared to the reference state (baseline category). For example, if the coefficient for Florida is negative, it means Trump polls lower in Florida compared to the reference state. The state-level effects account for regional differences in Trump’s support. Some states may show significantly higher or lower levels of support, even after adjusting for the time of the poll and pollster quality.
- Transparency Score ($\beta_4$): This coefficient shows how much Trump’s polling percentage is affected by the transparency of the poll. A positive coefficient would indicate that polls with higher transparency tend to report higher polling percentages for Trump, whereas a negative coefficient would imply the opposite.
- Days Until Election ($\beta_5$): The counting down days is a time-related variable. A positive and significant coefficient would suggest that Trump’s polling percentage has increased as the election date approaches, while a negative coefficient would suggest a decrease in his polling numbers over time.The coefficient for end_date informs us about how Trump’s polling percentages have evolved over time. A positive coefficient would suggest an upward trend, while a negative coefficient would indicate a decline.

## Model Justification
Based on the model summary shown in @tbl-summary-model, we observe that several state-level coefficients are statistically significant, indicating regional variations in support for Trump. Additionally, the coefficients for `transparency_score` and `days_until_election` are both highly significant, suggesting that pollster quality and the timing of the polls have notable effects on Trump’s polling percentages. A negative coefficient for `days_until_election` suggests a slight decline in support as the election approaches.

When evaluated on the test set, the model appears to struggle with accurately identifying Trump supporters, possibly due to unobserved factors or limitations in capturing the complexity of voter behavior.

The rationale for applying multiple regression in this context is to control for various influential factors simultaneously—such as time trends (`days_until_election`), pollster quality (`transparency_score`), and regional differences (`state-level effects`). This approach allows us to isolate the individual impact of each variable, providing a more detailed understanding of how each factor contributes to Trump’s polling outcomes. The full coefficient output is shown in the [Appendix -@sec-model-details]

```{r}
#| echo: false
#| label: tbl-summary-model
#| tbl-cap: Regression Model Summary Table
#| warning: false
#| message: false

regression_model<-readRDS(here::here("models/regression_model.rds"))
summary<-summary(regression_model)
# Extract the coefficients to form a table
coefficients_table <- summary$coefficients

# Display the first 10 rows of the coefficients table using kable, formatted to 2 decimal places
kable(head(coefficients_table, 10), digits = 2)
```

# Results {#sec-results}

@@

```{r}
#| echo: false
#| label: tbl-prediction
#| tbl-cap: Prediction for Trump and Harris by Electoral College
#| warning: false
#| message: false

# Read the test data from the .parquet file
analysis_data_test <- read_parquet(here::here("data/02-analysis_data/test_data.parquet"))
analysis_data_train<- read_parquet(here::here("data/02-analysis_data/train_data.parquet"))
# Filter out states in the test data that don't exist in the training data
analysis_data_test <- analysis_data_test %>%
 filter(state %in% unique(analysis_data_train$state))

@@

# Create the table using kable
kable(state_predictions[, c("state", "predicted_percent_trump", "electoral_votes", "winner")], 
      col.names = c("State", "Trump Predicted %", "Electoral Votes", "Winner"), 
      digits = 2)
```

## Predicted Electoral Outcomes by State
The following table (@tbl-missing) serves as a supplement to the predicted outcomes, filling in the gaps for states where the model couldn’t provide predictions. It allows for a complete view of electoral projections by assigning a likely outcome in each missing state based on historical data. This approach ensures every state is accounted for in the final electoral vote totals.

```{r}
#| echo: false
#| label: tbl-missing
#| tbl-cap: Missing States with Predicted Winner Based on Historical Lean
#| warning: false
#| message: false
# Create the data frame with missing states, electoral votes, and predicted winner based on historical lean
missing_states <- data.frame(
  state = c("Alabama", "Alaska", "Arkansas", "Connecticut", "Delaware", "Hawaii", "Idaho", 
            "Illinois", "Indiana", "Kansas", "Kentucky", "Louisiana", "Mississippi", "North Dakota", 
            "Oklahoma", "Oregon", "South Carolina", "Tennessee", "Utah", "Washington", 
            "West Virginia", "Wyoming", "District of Columbia"),
  Electoral_Votes = c(9, 3, 6, 7, 3, 4, 4, 19, 11, 6, 8, 8, 6, 3, 7, 8, 9, 11, 6, 12, 4, 3, 3),
  Prediction_Based_on_Historical_Lean = c("Trump", "Trump", "Trump", "Harris", "Harris", 
                                          "Harris", "Trump", "Harris", "Trump", "Trump", 
                                          "Trump", "Trump", "Trump", "Trump", "Trump", 
                                          "Harris", "Trump", "Trump", "Trump", "Harris", 
                                          "Trump", "Trump", "Harris")
)

# Display the table with kable
kable(missing_states, col.names = c("State", "Electoral Votes", "Prediction Based on Historical Lean"))
```

The following map (@fig-statemap) shows the predicted winner for each state in the 2024 U.S. Presidential Election, based on the regression model's predicted vote percentages for Trump and Harris. The predicted outcome in the regression model reflects the geographic voting patterns, with Trump winning in traditionally Republican-leaning states like Alabama, Missouri, and Wyoming, while Harris dominates in Democratic strongholds such as California, New York, and Illinois. However, key battleground states such as Florida and Arizona are predicted to favor Harris, potentially determining the overall election outcome.

```{r fig-statemap}
#| echo: false
#| label: fig-statemap
#| tbl-cap: Predicted Winner of 2024 US Presidential Election by State
#| warning: false
#| message: false

state_predictions <- merge(state_predictions, missing_states, by = "state", all = TRUE)

@@

The table below (@tbl-numstate) shows the numerical predicted winner of the 2024 US Presidential Election by state. Harris is expected to win the 2024 U.S. Presidential Election, securing 387 electoral votes, compared to Trump’s 151 electoral votes. The election outcome hinges on several battleground states, where the vote margins are predicted to be narrow.
```{r}
#| echo: false
#| label: tbl-numstate
#| tbl-cap: Numerical Predicted Winner of 2024 US Presidential Election by State
#| warning: false
#| message: false

@@

# Display the summary table in kable format
kable(summary_df, col.names = c("Candidate", "Total Electoral Votes"), format = "markdown")
```

@@

## Overview of Predictive Modeling for the 2024 U.S. Election
This paper employs a regression model to forecast the 2024 U.S. Presidential Election outcomes between Kamala Harris and Donald Trump. By analyzing historical state-level voting data, we estimated Trump’s vote percentages across states and calculated each candidate’s projected electoral votes. Our model predicts a significant victory for Harris, who is projected to secure 387 electoral votes compared to Trump’s 151. This finding underscores the impact of demographic and geographic factors on the U.S. electoral landscape. A geographic visualization further emphasizes the anticipated state-by-state outcomes and reveals enduring regional voting patterns.

## Regional Voting Patterns and Their Impact on Electoral Outcomes
Our analysis demonstrates the persistence of geographic voting patterns in U.S. elections. As expected, traditional Democratic strongholds such as California, New York, and Illinois favor Harris, while Republican-dominated states like Alabama and Wyoming show support for Trump. These regional preferences align with historical voting behaviors, suggesting that party loyalty and demographic factors continue to shape election results.

The model also highlights the critical role of battleground states, with states such as Florida and Arizona predicted to lean toward Harris. The influence of these states, where margins are often narrow, emphasizes the importance of campaign efforts and voter mobilization in these regions. The projections illustrate that minor shifts in voter sentiment within these states could decisively impact the overall election outcome.

## Polling and Electoral Uncertainty
The model’s reliance on historical polling data and state-level predictions reflects certain limitations inherent in election forecasting. For example, while the model captures general trends, it does not account for potential turnout variations, third-party candidates, or unexpected political events that might affect voter behavior. Furthermore, the use of polling data presents a challenge due to its variability; certain states with less frequent polling data show broader confidence intervals, indicating greater uncertainty in those regions.

This variability suggests that further studies could benefit from integrating real-time data and employing Bayesian methodologies to account for evolving voter preferences. Such an approach could improve the accuracy of predictions, especially in regions with historically unpredictable outcomes. Future research may also consider factors like demographic shifts and economic conditions, which could significantly influence voter preferences over time.

## Implications for Future Electoral Modeling
The findings from this analysis underscore the need for a nuanced approach to election forecasting. While historical voting patterns provide a solid foundation, they may not fully capture the complexities of modern elections, where factors such as social media influence and voter sentiment play increasingly prominent roles. Incorporating these additional variables, particularly in battleground states, could refine the model’s predictive power and provide a more comprehensive understanding of the electoral landscape.

For future iterations, we recommend expanding the model to include demographic data, such as age, education level, and income, which could reveal deeper insights into voter behavior. Additionally, real-time polling updates could offer a dynamic perspective, allowing the model to adapt to changes in public opinion as the election date approaches. Employing these enhancements would improve model reliability, particularly in closely contested states where the margin of error can significantly impact the final prediction.

## Weakness
Despite the model's strengths, it has limitations that requires further exploration. The assumption of linear relationships between predictors may oversimplify complex voter behaviors and social influences. The dataset simplifies the potentioanl influecne behind the US election. Additionally, the reliance on past data assumes that historical trends will persist, which may not hold true in the rapidly shifting political climate of the U.S. To address these issues, future models could benefit from incorporating voter turnout projections and real-time sentiment analysis, potentially drawing data from social media platforms and news sources to gauge public opinion. Moreover, a comparative post-election analysis could offer valuable insights, revealing areas where the model succeeded and where it may need refinement. This iterative approach would help create a more resilient and adaptable electoral forecasting model.

@@

# Appendix {#sec-appendix}

@@

### **Survey Structure**

@@

```{r}
#| echo: false
#| label: fig-diagnosis-1
#| tbl-cap: Linear regression assumptions hold.
#| warning: false
#| message: false
#| fig.width: 10
#| fig.height: 8

par(mfrow = c(2, 2))  
plot(regression_model)
```

## Calculate Mean Squared Error and Mean Absolute Error on Test Data
```{r}
#| echo: false
#| label: fig-diagnosis-2
#| tbl-cap: Calculate Mean Squared Error (MSE) on test data
#| warning: false
#| message: false

# Predicted and actual values
predicted_values <- predict(regression_model, newdata = analysis_data_test)
actual_values <- analysis_data_test$percent

# Calculate Mean Squared Error (MSE)
mse <- mean((predicted_values - actual_values)^2)

# Calculate Mean Absolute Error (MAE)
mae <- mean(abs(predicted_values - actual_values))

# Calculate R-squared
ss_total <- sum((actual_values - mean(actual_values))^2)
ss_residual <- sum((predicted_values - actual_values)^2)
r_squared <- 1 - (ss_residual / ss_total)

# Combine metrics into a data frame for display
metrics_df <- data.frame(
  Metric = c("Mean Squared Error (MSE)", "Mean Absolute Error (MAE)", "R-squared"),
  Value = c(mse, mae, r_squared))

# Display the metrics table using kable
kable(metrics_df, caption = "Model Performance Metrics on Test Data")
```

## Model Summary
```{r}
#| echo: false
#| tbl-cap: Complete Model Coefficient Summary Table
#| warning: false
#| message: false

# Get summary of the regression model
model_summary <- summary(regression_model)

# Extract and convert the coefficients summary to a data frame
coefficients_df <- as.data.frame(model_summary$coefficients)
colnames(coefficients_df) <- c("Estimate", "Std. Error", "t value", "Pr(>|t|)")

# Extract additional statistics
r_squared <- model_summary$r.squared
adj_r_squared <- model_summary$adj.r.squared
f_statistic <- model_summary$fstatistic[1]
p_value <- pf(f_statistic, model_summary$fstatistic[2], model_summary$fstatistic[3], lower.tail = FALSE)
residual_se <- model_summary$sigma

# Create a data frame for additional statistics with consistent column names
additional_stats <- data.frame(
  Estimate = c(r_squared, adj_r_squared, f_statistic, p_value, residual_se),
  `Std. Error` = NA,
  `t value` = NA,
  `Pr(>|t|)` = NA,
  row.names = c("R-squared", "Adjusted R-squared", "F-statistic", "F-statistic p-value", "Residual Std. Error")
)

# Ensure column names are identical between data frames
colnames(additional_stats) <- colnames(coefficients_df)

# Combine the coefficients and additional statistics
full_summary_df <- rbind(coefficients_df, additional_stats)

# Display the full summary table using kable
kable(full_summary_df, caption = "Regression Model Summary with Additional Statistics")
```

## Multicollinearity Check on the Training Data
```{r}
#| echo: false
#| warning: false
#| message: false

# Calculate the correlation matrix
cor_matrix <- cor(analysis_data_train[, c("numeric_grade", "sample_size", "transparency_score", "days_until_election")], use = "complete.obs")

# Rename the row and column names of the correlation matrix to remove underscores
colnames(cor_matrix) <- gsub("_", " ", colnames(cor_matrix))
rownames(cor_matrix) <- gsub("_", " ", rownames(cor_matrix))

# Print the correlation matrix
print(cor_matrix)

# Create a beautiful correlation plot with modified labels
corrplot(
  cor_matrix, 
  method = "color",      # Use color to fill the matrix
  type = "upper",        # Show only the upper triangle
  addCoef.col = "black", # Add correlation coefficients in black
  tl.col = "black",      # Label text color
  tl.srt = 45,           # Rotate labels
  number.cex = 0.8,      # Text size for coefficients
  col = colorRampPalette(c("red", "white", "blue"))(200) # Color gradient
)

@@
